---
title: 'Reproducible Research: Peer Assessment 1'
output:
  html_document:
    fig_height: 4
    fig_width: 5
    keep_md: yes
---

For this analysis we use the following libraries as dependencies and hence they have to be loaded into the environment.

```{r, results = 'hide', message = FALSE}
library(dplyr)
library(ggplot2)
```

## Loading and preprocessing the data

Load the data from the provided zip file:

```{r}
conn <- unz("activity.zip", filename = "activity.csv")
activity_data <- read.csv(conn, colClasses = c("integer", "Date", "integer"))
```

## What is mean total number of steps taken per day?

First, we want to summarize the data and compute the total number of steps made in each day of the measurement period. 

```{r}
daily_steps <- activity_data %>%
    group_by(date) %>% 
    summarise(steps = sum(steps, na.rm = T))
```

Next, let's take a look at the distribution of the daily number of steps:

```{r}
with(daily_steps, hist(steps, breaks=max(steps) / 1000))
```

And finally, let's see what are the mean and median total number of steps taken each day.

```{r}
with(daily_steps,
     list(mean = mean(steps),
          median = median(steps)))
```

## What is the average daily activity pattern?

In order to visualize the average daily activity pattern, we need to summarise the data and compute the average of steps taken within each time interval across all the days.

```{r}
activity_pattern <- activity_data %>%
    group_by(interval) %>% 
    summarise(avg_steps = mean(steps, na.rm = T))
```

Next, we need to plot the results.

```{r}
ggplot(mapping = aes(x = interval, y = avg_steps), data=activity_pattern) + 
    geom_line() +
    labs(title="Average steps taken per time interval", y = "average steps taken")
```

Here's an interesting question: which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r}
filter(activity_pattern, avg_steps == max(avg_steps))$interval
```

The result tells us that the most walking on average is done from 8:35 am to 8:40 am, which is consistent with the pattern of a person going to work in the morning, probably to start the work day around 9:00 am.

## Imputing missing values

Note that there are a number of days/intervals where there are missing values (coded as ```NA```). The presence of missing days may introduce bias into some calculations or summaries of the data.

Let's see how many observations with missing values for the step measurement are.

```{r}
summarise(activity_data, NAs = sum(is.na(steps)), NAs.percentile = NAs / n() * 100)
```

## Are there differences in activity patterns between weekdays and weekends?
